---
title: "Automobile transmission type and fuel mileage"
author: "Timo Voipio"
date: "21 Aug 2016"
output:
    pdf_document:
      latex_engine: xelatex
#mainfont: "texgyretermes-regular.otf"
#mathfont: "texgyretermes-math.otf"
fontsize: 12pt
classoption: a4paper
# This file is a WIP for using metric units (out of general interest);
# original question requires MPG
---

```{r echo=FALSE}
library(knitr)

# TRUE relegates most code chunks to appendices and disables result
# output by default
tidydoc <- FALSE

# figure width 4 inches (PDF output), height = width, center figures
opts_chunk$set(fig.width=4, fig.height=4, fig.align='center')

# Increase the penalty for using scientific notation (to format
# numbers like 10000 normally, not in scientific notation)
options(scipen = 3)

if (tidydoc)
{
    opts_chunk$set(results='hide', echo=FALSE, size='scriptsize')
}
```

# Executive summary

# Introduction

# Exploratory analysis

```{r init}

library(datasets)
data("mtcars")

library(ggplot2)
library(GGally)
```

The dataset includes the following variables:

Column name | Variable description and units
------------|-------------------------------
`mpg`       | Miles/(US) gallon (*gas mileage; 235 l/100 km = 1/(1 mpg)*)
`cyl`       | Number of cylinders
`disp`      | Displacement (cu.in.) (*1 litre ≈ 61.0 cu in*)
`hp`        | Gross horsepower (*1 kW ≈ 1.34 hp*)
`drat`      | Rear axle ratio (*driveshaft rpm/axle rpm*\footnote{"the drive-axle ratio is a comparison of the number of gear teeth on the ring gear of the rear axle and the pinion gear on the driveshaft. - - For example, a 4.11:1 ratio means there are 4.11 teeth on the axle's ring gear for each tooth on the driveshaft's pinion gear. Or, put another way, the driveshaft must turn 4.11 times to turn the rear wheels one full revolution. - - typical rule of thumb: The higher the numerical ratio, the slower the gear will be. This higher ratio gives a truck greater pulling power, but since the engine must work harder to spin the driveshaft more times for each turn of the rear wheels, top-end speed and fuel economy are sacrificed." Quoted from [here][drat]})
`wt`        | Weight (1000 lbs *≈ 454 kg*)
`qsec`      | 1/4 mile time
`vs`        | V/S [*V engine (0) or straight engine (1)*]
`am`        | Transmission (0 = automatic, 1 = manual)
`gear`      | Number of forward gears
`carb`      | Number of carburetors

```{r explore}

# Possibly relevant covariates:
# disp, hp, cyl (likely correlated with each other)
# wt
# drat 
# am (0 auto/1 manual; subject of interest)
# Outcome:
# mpg

# Convert transmission, cylinder count, carburetor count, and engine type
# to factor variables
mtcars$am <- factor(mtcars$am, levels = c(0, 1),
                    labels = c("auto", "manual"))
#mtcars$cyl <- factor(mtcars$cyl)
#mtcars$carb <- factor(mtcars$carb)
mtcars$vs <- factor(mtcars$vs, levels = c(0, 1),
                    labels = c("v", "straight"))

# Create a new data frame for metric variables
mtcars.si <- mtcars
mtcars.si$cons <- 235/mtcars$mpg # consumption in litres/100 km
mtcars.si$wt <- mtcars$wt*0.454 # weight in tonnes
mtcars.si$disp <- mtcars$disp/61.0 # displacement in litres
mtcars.si$pw <- mtcars$hp/1.34 # power in kilowatts
mtcars.si$mpg <- NULL
mtcars.si$hp <- NULL

# Create a new variable for gross power/weight ratio
# (kW/tonne)
mtcars.si$pwr <- mtcars.si$pw / mtcars.si$wt

fit1 <- lm(mpg ~ am, mtcars)
fitall <- lm(mpg ~ . - pwr, mtcars)

g <- ggplot(mtcars, aes(x = am, y = mpg))
g <- g + geom_boxplot() +
    stat_summary(fun.y = mean, geom = "point")
g <- ggtitle ("Gas mileage and transmission type")

print(g)

ggpairs(mtcars)

summary(fit1)


fit2.w <- lm(mpg ~ am + wt, mtcars)
fit2.d <- lm(mpg ~ am + disp, mtcars)
fit2.h <- lm(mpg ~ am + hp, mtcars) # greatest reduction in sum(^2)

anova(fit1, fit2.w)
anova(fit1, fit2.d)
anova(fit1, fit2.h)

fit3.w <- lm(mpg ~ am + hp + wt, mtcars) # greatest reduction in sum(^2)
fit3.d <- lm(mpg ~ am + hp + disp, mtcars)
fit3.c <- lm(mpg ~ am + hp + cyl, mtcars)

anova(fit1, fit2.h, fit3.w)
anova(fit1, fit2.h, fit3.d)
anova(fit1, fit2.h, fit3.c)

fit4.d <- lm(mpg ~ am + hp + wt + disp, mtcars)
fit4.c <- lm(mpg ~ am + hp + wt + cyl, mtcars) # smallest sum(^2), but fit4.c not really significant

anova(fit1, fit2.h, fit3.w, fit4.d)
anova(fit1, fit2.h, fit3.w, fit4.c)

fit5 <- lm(mpg ~ am + hp + wt + disp + cyl, mtcars)

anova(fit1, fit2.h, fit3.w, fit4.d, fit5)
anova(fit1, fit2.h, fit3.w, fit4.c, fit5) 

```

[drat]: http://www.worktruckonline.com/channel/vehicle-research/article/story/2011/07/10-factors-to-consider-when-spec-ing-drive-axle-ratios-for-medium-duty-trucks.aspx "10 Factors to Consider when Spec’ing Drive-Axle Ratios for Medium-Duty Trucks"
